---
title: "<center>Income</center>"
# subtitle: "Table of Contents"
output:
  html_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 8, fig.align = "center")

library(tidyverse)
library(scales)
library(plotly)

source("tools/authenticate.R")
source("tools/get_data.R")

# Authenticate user
authenticated <- google_authenticate(service_account_key = Sys.getenv("GARGLE_KEY_PATH"),
                                     gmail = Sys.getenv("MY_EMAIL"))

theme_update(plot.title = element_text(hjust = 0.5), 
             plot.subtitle = element_text(hjust = 0.5),
             panel.background = element_blank(),
             panel.grid.major = element_blank(),
             panel.grid.minor = element_blank(),
             axis.line = element_line(color = "black"))
```

All data in this report is **`r if_else(authenticated, "REAL", "SIMULATED")`**.

# Income by Month
There are various sources of income, ranging from paychecks to gifts to reimbursements (e.g. work travel, tax refund). Paychecks are recorded based on the net income after taxes, benefits, and other deductions. Complete data ranges back to approximately May 2025.
```{r income-data, include=FALSE}
income_sheet <- get_income_data(authenticated)
income <- format_date(income_sheet) %>% 
  filter(month > as.Date("2025-04-01")) %>% 
  mutate(source = as.factor(source))
```

```{r income-total}
income_total <- income %>% 
  group_by(year, month) %>% 
  summarize(amount = sum(amount), source = "Total", inputs = n(), date = max(date), 
            year = year(date), month = floor_date(date, unit = "month"),
            .groups = "drop") %>% 
  bind_rows(income)
income_total$source <- factor(income_total$source, 
                              levels = c(setdiff(levels(as.factor(income_total$source)),
                                                 c("Other", "Total")), 
                                         "Other", "Total"))

# Set income palette
income_palette <- hue_pal()(length(levels(income_total$source)) - 1)
```

<!--- The plot below shows income based on the date it was received (no time grouping). Note that the *y*-axis is on a logarithmic (base 10) scale to show greater detail at both lower income amounts (e.g. gifts, reimbursements) and higher amounts (paychecks). --->

```{r plot-income-total, eval=FALSE}
# Total not on monthly scale
income_plot <- income_total %>% 
  ggplot(aes(x = date, y = amount, col = source, group = source, 
             text = paste0("<b>", source, "</b><br><i>", date, 
                           "</i><br>Amount: ", dollar(amount)))) +
  geom_line() +
  geom_point(size = 0.5) +
  scale_y_log10() +
  scale_color_manual(values = c(income_palette[1], income_palette[2], income_palette[3],
                                income_palette[4], income_palette[5], income_palette[6],
                                income_palette[7], "black")) +
  labs(title = "Total Income",
       x = "Date",
       y = "Income ($)",
       col = "Income Source")
ggplotly(income_plot, tooltip = "text")
```

We will display income by month to better contextualize what is encompassed by the total. Due to biweekly (rather than monthly) paychecks, the total will fluctuate intermittently. The *y*-axis is on a logarithmic (base 10) scale to show greater detail at both lower income amounts (e.g. gifts, reimbursements) and higher amounts (paychecks).

```{r monthly-income, warning=FALSE, out.width="100%"}
income_month <- income %>% 
  group_by(year, month, source) %>% 
  summarize(amount = sum(amount), inputs = n(), .groups = "drop") %>% 
  bind_rows(filter(income_total, source == "Total"))

color_map <- c(income_palette[1], income_palette[2], income_palette[3],
               income_palette[4], income_palette[5], income_palette[6],
               income_palette[7], "black")

income_plot_month <- income_month %>% 
  ggplot(aes(x = month, y = amount, col = source, group = source, 
             text = paste0("<b>", source, "</b><br><i>", format(month, "%B %Y"), 
                           "</i><br>Amount: ", dollar(amount), "<br>Inputs: ", inputs))) +
  geom_line() +
  geom_point(size = 0.5) +
  scale_y_log10() +
  scale_color_manual(values = color_map) + 
  labs(title = "Total Income by Month",
       x = NULL,
       y = "Income ($)",
       col = "Income Source")
ggplotly(income_plot_month, tooltip = "text")
```

When comparing with total expenses, make sure to consider that insurance costs (medical, dental, and vision) have already been deducted from the income recorded.
